---
title: "differential-expression-analysis"
author: "Victoria Latynina"
date: "2023-03-21"
output:
  pdf_document: default
  html_document: default
---

## Differential Expression analysis

Dataset: GSE106542 (<https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE106542>)

Differences in expression between TEMRA IL7-high and TEMRA IL7 low

### Pipeline steps:

-   Preprocess counts data

-   Run VST for PCA, remove outliers, run PCA

-   Run differential expression

-   Find pathways for DE

```{r rna, warning=F, message=F}
library(ggplot2)
library(DESeq2)
library(apeglm)
library(ggrepel)
library(dplyr)
library(org.Hs.eg.db)
library(PCAtools)
library(GEOquery)
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=40),tidy=TRUE)
```

## Preprocess counts data

```{r rna1, warning=F, message=F, tidy=TRUE}
path <- "~/Code/sys_bio/Part_2/materials/GSE106542_RAW/GSE106542_Bulk_raw_counts.txt"
data <- read.table(path, row.names = 1, sep="\t", header = 1)
data$gene_symbol <- mapIds(org.Hs.eg.db, 
                           gsub("\\.\\d+", "", rownames(data)), 
                           column="SYMBOL", "ENSEMBL")
data$gene_name <- mapIds(org.Hs.eg.db, 
                         gsub("\\.\\d+", "", rownames(data)), 
                         column="GENENAME", "ENSEMBL")

mapping <- data[, c("gene_symbol", "gene_name"), 
                drop = TRUE]
counts <- data[, 2:ncol(data)-2]
colnames(counts) <- gsub("BRNA_", "", colnames(counts))

annotations <- getGEO("GSE106542")[[1]]
pdata <- pData(annotations)[, c("title", 
                                "description", "organism_ch1", 
                                "molecule_ch1", "subject #:ch1", 
                                "cell subtype surface markers:ch1", 
                                "cell subtype:ch1", "cell type:ch1", 
                                "longitudinal visit:ch1")]
rownames(pdata) <- gsub("Bulk_RNA-seq_", "", pdata$title)
```

### Match columns and rows for all the data

```{r rna129, warning=F, message=F, tidy=TRUE}
col_order <- rownames(pdata)
counts <- counts[, col_order]
identical(colnames(counts), rownames(pdata))
```

### Create a new dataset without "TEM" and "TCM" rows

```{r rna21, warning=F, message=F, tidy=TRUE}
temra_pdata <- subset(pdata, 
pdata$`cell subtype:ch1` != "TEM" & pdata$`cell subtype:ch1` != "TCM")
col_order <- rownames(temra_pdata)
temra_counts <- counts[, col_order]
identical(colnames(temra_counts), 
          rownames(temra_pdata))
```

## Run differential expression for all cell subtypes

```{r rna22, warning=F, message=F, tidy=TRUE}
pdata$Cell_subtype <- as.character(pdata$`cell subtype:ch1`)
pdata$Cell_subtype <- factor(pdata$Cell_subtype, 
                             levels=c("TEM", 
                                      "TCM", 
                                      "IL7Rhigh_TEMRA", 
                                      "IL7Rlow_TEMRA"))
pdata$Donor <- as.character(pdata$`subject #:ch1`)
pdata$Donor <- factor(pdata$Donor, 
                             levels=c("Subject16", 
                                      "Subject20", 
                                      "Subject21", 
                                      "Subject22",
                                      "Subject23"))
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = pdata,
                              design= ~ Cell_subtype + Donor)
dds <- DESeq(dds)
resultsNames(dds)
```

## Run VST for PCA, run PCA and notice outliers

```{r rna33, warning=F, message=F, tidy=TRUE}
vst <- varianceStabilizingTransformation(dds)
plotPCA(vst, 
        intgroup=c("Cell_subtype"), 
        ntop=nrow(vst)) + theme_bw() +
  theme(aspect.ratio = 1)

```

## Add thresholds to get rid of outliers, plot PCA by donors and by cell subtypes

```{r rna310, warning=F, message=F, tidy=TRUE}
PCA_data <- plotPCA(vst, 
        intgroup=c("Cell_subtype", "Donor"), 
        ntop=nrow(vst), returnData = TRUE)

cell_colors <- c("TEM" = "red", "TCM" = "blue", "IL7Rhigh_TEMRA" = "green", "IL7Rlow_TEMRA" = "pink")
donor_shapes <- c("Subject16"=20, 
                  "Subject20"=21, 
                  "Subject21"=22, 
                  "Subject22"=23,
                  "Subject23"=24)
outliers <- PCA_data$PC2 > 5
PCA_data <- subset(PCA_data, !outliers)
ggplot(PCA_data, aes(x = PC1, y = PC2, color = Cell_subtype, shape = Donor)) + 
  geom_point(size = 4) + 
  theme_bw() +
  theme(aspect.ratio = 1) +
  scale_color_manual(values = cell_colors) +
  scale_shape_manual(values = donor_shapes)
```

```{r rna444, message=FALSE, warning=FALSE, echo=F, tidy=TRUE}
pca_data <- prcomp(t(assay(vst)))
pca_df <- data.frame(PC1 = pca_data$x[,1], 
                     PC2 = pca_data$x[,2],
                     Donor = colData(vst)$Donor)

threshold_PC1 <- 3 * sd(pca_df$PC1)
threshold_PC2 <- 0.38 * sd(pca_df$PC2)
fpca_df <- pca_df[
  abs(pca_df$PC1) < threshold_PC1 & abs(pca_df$PC2) < threshold_PC2,]

p <- ggplot(fpca_df, aes(x = PC1, y = PC2, color = Donor)) +
  geom_point() +
  theme_bw() +
  theme(aspect.ratio = 1)
#print(p)
```

```{r rna44, warning=F, message=F, echo=F, tidy=TRUE}
pca_data <- prcomp(t(assay(vst)))
pca_df <- data.frame(PC1 = pca_data$x[,1], 
                     PC2 = pca_data$x[,2],
                     Cell_subtype = colData(vst)$Cell_subtype)

threshold_PC1 <- 3 * sd(pca_df$PC1)
threshold_PC2 <- 0.38 * sd(pca_df$PC2)
fpca_df <- pca_df[
  abs(pca_df$PC1) < threshold_PC1 & abs(pca_df$PC2) < threshold_PC2,]

p <- ggplot(fpca_df, aes(x = PC1, y = PC2, color = Cell_subtype)) +
  geom_point() +
  theme_bw() +
  theme(aspect.ratio = 1)
# print(p)
```

## Run differential expression for target cell subtypes ("IL7Rhigh_TEMRA", "IL7Rlow_TEMRA")

```{r rna2, warning=F, message=F, tidy=TRUE}
temra_pdata$Cell_subtype <- as.character(temra_pdata$`cell subtype:ch1`)
temra_pdata$Cell_subtype <- factor(temra_pdata$Cell_subtype, levels=c("IL7Rhigh_TEMRA", "IL7Rlow_TEMRA"))
temra_pdata$Donor <- as.character(temra_pdata$`subject #:ch1`)
temra_pdata$Donor <- factor(temra_pdata$Donor, 
                             levels=c("Subject16", 
                                      "Subject20", 
                                      "Subject21", 
                                      "Subject22",
                                      "Subject23"))
dds <- DESeqDataSetFromMatrix(countData = temra_counts,
                              colData = temra_pdata,
                              design= ~ Cell_subtype + Donor)
dds <- DESeq(dds)
resultsNames(dds)
```

## PCA of target cell subtypes

```{r, rna222, warning=F, message=F, tidy=TRUE}
vst <- varianceStabilizingTransformation(dds)

pca_data <- prcomp(t(assay(vst)))
pca_df <- data.frame(PC1 = pca_data$x[,1], 
                     PC2 = pca_data$x[,2],
                     Cell_subtype = colData(vst)$Cell_subtype,
                     Donor = colData(vst)$Donor)

threshold_PC1 <- 80 #50 * sd(pca_df$PC1)
threshold_PC2 <- 10 * sd(pca_df$PC2)
fpca_df <- pca_df[
  abs(pca_df$PC1) < threshold_PC1 & abs(pca_df$PC2) < threshold_PC2,]

ggplot(fpca_df, aes(x = PC1, y = PC2, color = Cell_subtype, shape = Donor)) + 
  geom_point(size = 4) + 
  theme_bw() +
  theme(aspect.ratio = 1) +
  scale_color_manual(values = cell_colors) +
  scale_shape_manual(values = donor_shapes)
```

## Get volcano plots

```{r rna4, warning=F, message=F, tidy=TRUE}
res <- lfcShrink(dds, coef="Cell_subtype_IL7Rlow_TEMRA_vs_IL7Rhigh_TEMRA", 
                 type="apeglm")
res$gene_symbol <- mapIds(org.Hs.eg.db, 
                          gsub("\\.\\d+", "", rownames(res)), 
                          column="SYMBOL", "ENSEMBL")
head(res)
resDF <- as.data.frame(res)
ggplot(resDF, aes(x=log2FoldChange, y=-log10(padj), color=padj < 0.05)) +
  geom_point() + theme_bw() + scale_color_manual(values=c("black", "red")) +
  geom_text_repel(data=resDF %>% dplyr::filter(padj < 1e-7), 
aes(label=gene_symbol), color="black") +
xlim(c(-8, 8)) + 
  xlab("low TEMRA    Log2FC    high TEMRA") + 
  geom_vline(xintercept = 0, lty=2) +
  theme(aspect.ratio = 1)
```

## Pathway analysis

```{r rna5, echo=T, warning=F, message=F, tidy=TRUE}
library(fgsea)

deResults <- results(dds)
deResults$gene_symbol <- mapIds(org.Hs.eg.db, 
                                gsub("\\.\\d+", "", rownames(deResults)), 
                                column="SYMBOL", "ENSEMBL")
stats <- deResults$stat
names(stats) <- deResults$gene_symbol
complete_cases <- complete.cases(stats)
stats <- stats[complete_cases]
top_genes <- resDF %>% dplyr::filter(padj < 1e-7)
```

```{r rna6, echo=T, warning=F, message=F, tidy=TRUE}
load("~/Code/sys_bio/Part_2/materials/keggSymbolHuman.rdata")
fgseaResults <- fgseaMultilevel(keggSymbolHuman, 
                                stats, 
                                minSize = 15, 
                                maxSize = 500)

topPathwaysUp <- fgseaResults[ES > 0, ][head(order(pval), n=8), pathway]
topPathwaysDown <- fgseaResults[ES < 0, ][head(order(pval), n=8), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(keggSymbolHuman[topPathways], stats, fgseaResults, gseaParam = 0.5,
              pathwayLabelStyle=list(size=6))
```
